# Welcome and Set Up

These are the course notes and not a textbook. They are more like a field guide written by someone who has gotten lost in these woods before and wants to help you find your way through.

The notes are hands-on by design. You will write code, look at data, make maps, fit models, and interpret results. Each section builds on what came before. The math shows up when it needs to, but the emphasis throughout is on doing spatial analysis and understanding what you're doing -- not on deriving things for their own sake.

I'll point you to textbooks and papers along the way. Read them. They complement what we're doing here and will matter when you're working on your own data and I'm not around to ask.

These notes are a living document. They change based on what works, what doesn't, and what you ask about in class. If something is wrong or unclear, tell me.

## How These Notes Are Laid Out

Most of the chapters follow a natural progression -- each one builds on the last, and working through them in order is the right call. But scattered through the notes you'll find **Asides**: short detours that dig into something that doesn't quite fit the main flow. An Aside might work through the math behind a method, explain how R handles something under the hood, or fill in background that makes the surrounding chapters make more sense.

You can skip the Asides and still follow the core material. But they're there because the questions they answer are real ones -- the kind that come up when you're working through a chapter and start wondering why something works the way it does. If that's how your brain works, the Asides are for you.

## What's in These Notes

The notes move through the core ideas of spatial analysis in roughly this order:

- **Introduction: Getting Your Bearings** -- Before we can analyze spatial data, we need to be able to handle it. This chapter gets you oriented with the tools and vocabulary we'll use throughout.
- **Aside: Methods and Generics in R** -- A quick look under the hood at how R uses generic functions. Knowing this will help you make sense of why `plot()` and `summary()` behave differently depending on what you give them.
- **Mt Baker Map** -- A practical introduction to raster and vector spatial data in R, using elevation and infrastructure data from a local ski area.
- **Aside: Distance Matrices** -- A detour into how distances between points are computed and stored. This comes up enough in spatial analysis that it's worth understanding before you need it.
- **Point Patterns** -- How to quantify whether points in space are randomly distributed, clustered, or repulsed, using kernel density estimation and Ripley's K.
- **Aside: Correlation and Regression Are Not the Same Thing** -- These two ideas get conflated constantly. This aside untangles them before we lean on regression in later chapters.
- **Intro to Autocorrelation** -- How to measure and describe spatial structure in a continuous variable using variograms, Moran's I, and correlograms.
- **Geostats I: Inverse Distance Weighting** -- Predicting values at unsampled locations using a deterministic, distance-weighted approach. Simple, intuitive, and a good place to start.
- **Geostats I: Thin-Plate Splines** -- A smooth, flexible interpolation method that minimizes curvature across the surface.
- **Geostats II: Kriging** -- The probabilistic workhorse of geostatistics. Kriging uses the spatial structure of the data to predict unknown values and -- unlike IDW -- gives you a real measure of uncertainty.
- **Geostats II: Regression Kriging** -- Kriging with external predictors folded in. Useful when you have covariates that explain part of the spatial pattern.
- **Aside: Kriging Variance By Hand** -- Cracks open the `gstat` black box. We calculate kriging weights and prediction variance using matrix algebra on a toy dataset and check our work against the package.
- **Aside: OLS via Algebra and Matrices** -- Derives the OLS solution algebraically and in matrix form, then implements both in R. Worth reading before the GLS chapter, where GLS turns out to be OLS with one extra matrix in the mix.
- **Regression: GLS with Autocorrelated Residuals** -- What to do when your regression residuals have spatial structure. Spoiler: you use GLS, and it works.

## Go Do Great Things

You are going to learn to see the world spatially. That sounds dramatic, but I mean it practically: by the end of this course you will be able to look at a dataset, ask where things are happening and why, and have real tools to answer those questions. That is not a small thing.

The people who do this work well are not necessarily the ones who find it easiest. They're the ones who run the code when it breaks, read the error messages, ask questions, and keep going. That's it. That's the whole secret.

So: set up your project, download the data, and let's get to work.

## Technical Setup

This document was written in Markdown using using [quarto](https://quarto.org) and built with **R version `r getRversion()`**. You should be reasonably up to date on your versions of R, RStudio, and relevant packages. You can update your packages by running:

``` r
update.packages()
```

**Run that now.** And anytime it occurs to you. It's always a good idea to be up to date with your packages. 

### Project Structure

To follow along with the examples, you'll want a working RStudio project.

1.  **Create a new RStudio project**\
Go to **File → New Project → New Directory → New Project**. Give it a name (for example `spatial-course`) and choose where to save it.

2.  **Download the `data/` folder**\
The datasets used in the examples are available in the `data/` folder of the course repository. Download that folder and place it inside your project directory.

You can [download the data directly](https://github.com/AndyBunn/spatialNotes/releases/download/data-latest/data.zip) from the GitHub [repo](https://github.com/AndyBunn/spatialNotes). Here is hard link to it:

```
https://github.com/AndyBunn/spatialNotes/releases/download/data-latest/data.zip
```

Once it's unzipped your folder structure should look something like this:

   ```
   spatial-course/
   ├── data/
   │   ├── birdRichnessMexico.rds
   │   ├── mtbDEM.tif
   │   └── ...
   └── spatial-course.Rproj
   ```

3.  **Refer to data files using relative paths**

In your code, use paths like `"data/birdRichnessMexico.rds"` rather than full file paths. This keeps the code portable and ensures it will run on different machines without modification. E.g.,

```{r}
birdRichness <-  readRDS("data/birdRichnessMexico.rds")
```


4. **Save your Rmd files in the project root**

When you make a Rmd file for your work, save it in the project's root directory. For instance if you have the first two assignments saved in files called "MtBakerHomework.Rmd" and "pointPatternHomework.Rmd," your file and folder structure might look something like this:

   ```
   spatial-course/
   ├── data/
   │   ├── birdRichnessMexico.rds
   │   ├── mtbDEM.tif
   │   └── ...
   ├── MtBakerHomework.Rmd
   ├── pointPatternHomework.Rmd
   └── spatial-course.Rproj
   ```

```{r}
#| include: false
knitr::write_bib(c(
  .packages(), 'tidyverse', 'terra', 'sf', 'ggnewscale', 'tidyterra',
  'spatstat', 'ncf', 'spdep', 'gstat', 'tmap', 'nlme', 'fields', 'ggrepel'),
  'packages.bib')
# quarto render
```
