# Aside: Understanding Methods and Generics in R {- .aside-chapter}

## You’ve Come a Long Way in R

In your journey through the curriculum, you’ve learned a lot of R. You can wrangle data, fit models, make plots, write functions, etc. But you might also be starting to wonder:  **What’s actually going on under the hood?**

For example, how does `summary()` know what to do when you give it the output from `lm`? Or why does `plot()` sometimes give you a histogram and other times a scatterplot?

A good way to start answering those questions is to learn about **methods** and **generic functions**.

## Why Use Generics?

It’s nice when a function does the *right thing* depending on the kind of object you give it. You don’t want to memorize different function names for every kind of model—you just want to call `summary()`, `plot()`, or `print()` and get something useful. That’s what **generic functions** are for.

In R, many common functions like `summary()`, `plot()`, and `print()` are *generic*—they automatically call the version that matches the class of your object.

Let's run a linear model and use that as an example.

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
```

Ok. We have a new object called `fit` that is a `list` with a bunch of stuff in it that authors of the `lm` function think will be useful to you. Look at its structure to see what's in there.

```{r}
str(fit)
```

That list has 12 things in it including the coefficients, residuals, fitted values and a lot more. But getting the information you want in terms of interpreting the linear model isn't obvious. For instance, the t-statistics, p-values, and $R^2$ aren't sitting there waiting for you — `summary()` computes them on the fly from what *is* stored. That's part of what `summary.lm` actually does.

```{r}
summary(fit)
```

Ok. Check that out. We got a really nice formatted output that helps us interpret the linear model using the `summary` function. But note that the output of `summary(fit)` is really different than getting a summary of a numeric vector like the `mpg` in the `mtcars` data that went into the model.

```{r}
summary(mtcars$mpg)
```

### Same Function, Different Behavior. What Gives?

We used the same function (`summary`) and got two very different results.  
Answer? **Generics and methods.**

When we used `summary(fit)`, we're not calling some one-size-fits-all summary function. We're calling a **specific method** written for objects of class `"lm"`.

## How This Works

Here’s what R does when you call `summary(fit)`:

1. It checks the class of the object (`class(fit)`).
2. It looks for a function named `summary.classname`—in this case, `summary.lm`.
3. If it finds that method, it runs it.
4. If not, it falls back to `summary.default()`.

This system is called S3. It's the oldest and most common object system in R, and it's what you'll encounter in almost every package you use in this course.

You can look under the hood:

```{r}
class(fit)
```

That tells you `fit` is of class `"lm"`. Now you can see all the methods that currently exist for summary in your environment. If you load more packages and you'll see more methods show up here.

```{r}
methods(summary)
```

And if you want to see the actual code that runs when you call `summary(fit)`, use `getS3method`:

```{r}
getS3method("summary", "lm")
```

## What’s the Difference Between a Generic and a Method?

- A **generic function** is the function you call—like `summary()` or `plot()`.  
  It doesn’t actually do the work itself. Instead, it decides **which method to run** based on the class of the object.

- A **method** is the specific function written for a particular class.  
  For example:
  - `summary.lm()` handles linear models.
  - `summary.data.frame()` handles data frames.
  - `summary.default()` is the fallback if no class-specific method is found.

You can look at the generic directly:

```{r}
summary
```

That UseMethod("summary") in the function body is what triggers dispatch. That is, it tells R to go find the right method.

::: {#fig-s3dispatch}

```{=html}
<svg width="100%" viewBox="0 0 680 500" role="img" xmlns="http://www.w3.org/2000/svg">
<title>Generics and methods: S3 dispatch in R</title>
<desc>Three blob characters representing R objects with classes lm, data.frame, and numeric send arrows to the summary() generic function box in the center, which then dispatches to summary.lm(), summary.data.frame(), or summary.default() on the right.</desc>
<defs>
  <marker id="arrow" viewBox="0 0 10 10" refX="8" refY="5" markerWidth="6" markerHeight="6" orient="auto-start-reverse">
    <path d="M2 1L8 5L2 9" fill="none" stroke="context-stroke" stroke-width="1.5" stroke-linecap="round" stroke-linejoin="round"/>
  </marker>
</defs>

<text x="340" y="36" text-anchor="middle" font-family="sans-serif" font-size="15" font-weight="500" fill="#2d2d2a">generics and methods: how R dispatches your function call</text>

<text x="90" y="60" text-anchor="middle" font-family="sans-serif" font-size="12" fill="#888780">objects</text>
<text x="552" y="60" text-anchor="middle" font-family="sans-serif" font-size="12" fill="#888780">methods</text>

<g transform="translate(90,140)">
  <path d="M0,-44 C32,-50 58,-18 52,10 C54,36 30,50 0,46 C-30,50 -54,36 -52,10 C-58,-18 -32,-50 0,-44" fill="#FAECE7" stroke="#D85A30" stroke-width="1.2"/>
  <circle cx="-15" cy="-6" r="5.5" fill="#993C1D"/>
  <circle cx="15" cy="-6" r="5.5" fill="#993C1D"/>
  <circle cx="-13" cy="-7" r="2.5" fill="white"/>
  <circle cx="17" cy="-7" r="2.5" fill="white"/>
  <rect x="-20" y="14" width="40" height="19" rx="5" fill="#D85A30"/>
  <text x="0" y="27" text-anchor="middle" font-family="monospace" font-size="11" fill="white">"lm"</text>
</g>

<g transform="translate(90,278)">
  <path d="M0,-44 C32,-50 58,-18 52,10 C54,36 30,50 0,46 C-30,50 -54,36 -52,10 C-58,-18 -32,-50 0,-44" fill="#FAECE7" stroke="#D85A30" stroke-width="1.2"/>
  <circle cx="-15" cy="-6" r="5.5" fill="#993C1D"/>
  <circle cx="15" cy="-6" r="5.5" fill="#993C1D"/>
  <circle cx="-13" cy="-7" r="2.5" fill="white"/>
  <circle cx="17" cy="-7" r="2.5" fill="white"/>
  <rect x="-44" y="14" width="88" height="19" rx="5" fill="#D85A30"/>
  <text x="0" y="27" text-anchor="middle" font-family="monospace" font-size="11" fill="white">"data.frame"</text>
</g>

<g transform="translate(90,416)">
  <path d="M0,-44 C32,-50 58,-18 52,10 C54,36 30,50 0,46 C-30,50 -54,36 -52,10 C-58,-18 -32,-50 0,-44" fill="#FAECE7" stroke="#D85A30" stroke-width="1.2"/>
  <circle cx="-15" cy="-6" r="5.5" fill="#993C1D"/>
  <circle cx="15" cy="-6" r="5.5" fill="#993C1D"/>
  <circle cx="-13" cy="-7" r="2.5" fill="white"/>
  <circle cx="17" cy="-7" r="2.5" fill="white"/>
  <rect x="-36" y="14" width="72" height="19" rx="5" fill="#D85A30"/>
  <text x="0" y="27" text-anchor="middle" font-family="monospace" font-size="11" fill="white">"numeric"</text>
</g>

<rect x="278" y="218" width="124" height="100" rx="14" fill="#E1F5EE" stroke="#1D9E75" stroke-width="1.5"/>
<text x="340" y="241" text-anchor="middle" font-family="monospace" font-size="14" font-weight="500" fill="#085041">summary()</text>
<circle cx="325" cy="257" r="4" fill="#085041"/>
<circle cx="355" cy="257" r="4" fill="#085041"/>
<circle cx="327" cy="256" r="1.8" fill="white"/>
<circle cx="357" cy="256" r="1.8" fill="white"/>
<path d="M328,268 Q340,278 352,268" fill="none" stroke="#085041" stroke-width="1.5" stroke-linecap="round"/>
<text x="340" y="291" text-anchor="middle" font-family="sans-serif" font-size="11" fill="#0F6E56">generic function</text>
<text x="340" y="306" text-anchor="middle" font-family="sans-serif" font-size="11" fill="#0F6E56">dispatches by class &#x2192;</text>

<path d="M143,140 Q210,186 278,240" fill="none" stroke="#B4B2A9" stroke-width="1.3" marker-end="url(#arrow)"/>
<line x1="143" y1="278" x2="278" y2="278" stroke="#B4B2A9" stroke-width="1.3" marker-end="url(#arrow)"/>
<path d="M143,416 Q210,362 278,314" fill="none" stroke="#B4B2A9" stroke-width="1.3" marker-end="url(#arrow)"/>

<path d="M400,218 Q432,138 460,140" fill="none" stroke="#7F77DD" stroke-width="1.3" marker-end="url(#arrow)"/>
<line x1="402" y1="268" x2="460" y2="268" stroke="#7F77DD" stroke-width="1.3" marker-end="url(#arrow)"/>
<path d="M400,318 Q432,418 460,416" fill="none" stroke="#7F77DD" stroke-width="1.3" marker-end="url(#arrow)"/>

<rect x="460" y="118" width="185" height="44" rx="8" fill="#EEEDFE" stroke="#7F77DD" stroke-width="1.2"/>
<text x="552" y="140" text-anchor="middle" dominant-baseline="central" font-family="monospace" font-size="13" font-weight="500" fill="#3C3489">summary.lm()</text>

<rect x="460" y="246" width="185" height="44" rx="8" fill="#EEEDFE" stroke="#7F77DD" stroke-width="1.2"/>
<text x="552" y="268" text-anchor="middle" dominant-baseline="central" font-family="monospace" font-size="11" font-weight="500" fill="#3C3489">summary.data.frame()</text>

<rect x="460" y="394" width="185" height="44" rx="8" fill="#EEEDFE" stroke="#7F77DD" stroke-width="1.2"/>
<text x="552" y="416" text-anchor="middle" dominant-baseline="central" font-family="monospace" font-size="13" font-weight="500" fill="#3C3489">summary.default()</text>

<text x="552" y="453" text-anchor="middle" font-family="sans-serif" font-size="11" fill="#888780">&#x2191; fallback if no matching method</text>
</svg>
```

When you call a generic like `summary()`, R checks the class of the object and routes to the matching method. Load more packages and more methods show up. No class-specific method? R falls back to `summary.default()`.
:::


## Common Generic Functions

Here are some other generic functions you're likely to run into:

- `print()` – display a human-readable version  
- `summary()` – provide a richer summary  
- `plot()` – show plots  
- `predict()` – make predictions  
- `residuals()` – return residuals  
- `fitted()` – return fitted values  
- `coef()` – extract coefficients  
- `update()` – refit a model with changes  
- `AIC()` / `BIC()` – compare model fit  
- `anova()` – compare nested models  
- `terms()` / `formula()` – extract parts of model structure  

## How to Explore and Understand Methods

Want to know what’s going on behind a function call? Here are a few tools:

```{r}
methods(summary)
methods(plot)
methods(class = "lm")
getS3method("summary", "lm")
class(fit)
```

## Why This Matters

Understanding how generics and methods work helps you read other people's code, write better functions, and make sense of R's behavior when things feel a little magic.

You'll see this pattern come up throughout the course. When we work with spatial objects — things like `sf` features or `ppp` point patterns — the same idea applies. `plot()`, `summary()`, `print()` all do the right thing because someone wrote a method for that class. Now you know how.
