# Deterministic Interpolation: Inverse Distance Weighting

```{r,echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
set.seed(184)
knitr::purl("05aGeostatsIDW.qmd", output = "rScripts/05aGeostatsIDW.R", documentation = 1)

```

## Reading
Start with Hijmans's [Interpolation](https://rspatial.org/analysis/4-interpolation.html#). The examples using inverse weights to interpolate the California precipitation data are excellent. We will finish Ch 8 in Bivand next module so you can tackle that as time and motivation allow [@bivand2008geostatistics].

## Big Idea
We often want to estimate phenomena at locations where we do not have measurements. Geostatistics is the field that we use to make (hopefully) accurate and reliable models of spatial data. We will use these models for prediction. We can use both deterministic and probabilistic models and this module will focus on deterministic methods.

## Packages
The `gstat`[@R-gstat] package is your friend. And I'm going to plot in `ggplot` via `tidyverse`[@R-tidyverse] with `sf`[@R-sf], `terra`[@R-terra] and `tidyterra`[@R-tidyterra]. 


```{r, message=FALSE}
library(tidyverse)
library(sf)
library(gstat)
library(terra)
library(tidyterra)
```


## Geostats
The term "geostatistics" is big and vague -- like "statistics" it covers both descriptive and inferential approaches. The difference, somewhat tautologically, is the "geo": we use the spatial structure in the data to aid our understanding.

So far we've been mostly describing spatial patterns and sometimes using those patterns to make inferences about processes. Now we're going to predict. Collecting data is expensive and time consuming, so we often sample and then hope to say something about unsampled locations. In geostatistics, because we know how to measure spatial structure (autocorrelation, remember?), we can use that structure to produce predictions for locations we never visited. And with probabilistic methods, we get uncertainty estimates on those predictions too. We'll come back to that last part in the later chapters.

## Theory
We are going to tackle the broader theory and application of geostatistics (which did indeed develop from the desire to get more gold out of the ground in South Africa) but we are going to start with a conceptual application that predates geostatistics as a discipline. We have already introduced the idea of inverse distance weights (IDW) with our matrix $W$ from the Moran's I calculation where $w_{ij}=d_{ij}^{-1}$.

Using distance as a predictor implements the so-called first law of the geography "Everything is related to everything else, but near things are more related than distant things." IDW uses this assumption to predict a variable at unmeasured locations by using nearby measurements. The closer the measured values are to the prediction location, the more influence (weight) they have. With IDW, influence diminishes as a function of distance.

We are now going to use IDW as a way to get an estimate of a variable ($\hat{Z}$) at a spatial location $s_0$ using a weighted spatial average:

$$\hat{Z}(s_0)=\frac{\sum_{i=1}^n w(s_i)Z(s_i)}{\sum_{i=1}^n w(s_i)}$$

where $w$ is the weights of $n$ measurements at $i$ locations ($s$). The weight is calculated as a function of inverse distance:
$$ w(s_i) = ||s_i - s_0||^{-p} $$

where $|| \cdot ||$ indicates Euclidean distance and $p$ is a weighting power (often defaulting to two). The inverse power determines the influence that nearby points have ($w$ closer to one) as compared to points farther away ($w$ closer to zero) in the weighted average.

Let's visualize the function where weight ($w$) equals the inverse of distance ($d$): $w=d^{-1}$ as such:
```{r}
foo <- data.frame(d = 1:100, w = (1:100)^-1)
foo %>% ggplot(mapping = aes(x=d,y=w)) + geom_line() +
  labs(x="Distance",y="Weight")
```

And with a log scale to see more clearly.

```{r}
foo %>% ggplot(mapping = aes(x=d,y=w)) + geom_line() +
  labs(x="Distance",y="Weight") +
  scale_y_log10()
```

In that case we are using the straight inverse distance as $w=d^{-1}$ or $w=d^{-p}$ where $p=1$. But what if that curve is too steep? Or not steep enough? We can choose a value of $p$. E.g., the function can be $w=d^{-2}$ if we set $p=2$. Let's look at a range of $p$ values.
```{r}
foo <- rbind(data.frame(d = 1:100, w = (1:100)^0, p = "0"),
             data.frame(d = 1:100, w = (1:100)^-0.5, p = "0.5"),
             data.frame(d = 1:100, w = (1:100)^-1, p = "1"),
             data.frame(d = 1:100, w = (1:100)^-1.5, p = "1.5"),
             data.frame(d = 1:100, w = (1:100)^-2, p = "2"),
             data.frame(d = 1:100, w = (1:100)^-2.5, p = "2.5"))
foo %>% ggplot(mapping = aes(x=d,y=w,color=p)) + geom_line() +
  labs(x="Distance",y="Weight")
```

And with a log scale to see more clearly (if you can think in logs -- it takes me awhile to wrap my head around it, but then I'm glad I did).

```{r}
foo %>% ggplot(mapping = aes(x=d,y=w,color=p)) + geom_line() +
  labs(x="Distance",y="Weight") +
  scale_y_log10()
```

So, $p=0$ would weight all distances equally and as $p$ increases, the weights fall off quickly with distance.

## Toy example
### Data
Now that we have the information we need on the algorithm, let's do a quick and dirty example by hand to walk through some of the calculations. 

Here is a small data set of five points. We have spatial coordinates ($x$ and $y$) and some measured variable ($z$).
```{r}
foo <- data.frame(x = c(1,3,1,4,5),
                  y = c(5,4,3,5,1),
                  z = c(100,105,105,100,115))
foo
p1 <- foo %>% ggplot() + 
  geom_point(aes(x=x,y=y,size=z)) +
  lims(x=c(0,6),y=c(0,6))
p1
```

Let's imagine a point $x=2$ and $y=4$ where we don't know $z$.

```{r}
p1 + geom_point(aes(x=2,y=4),color="red",size=10,shape=0) +
  geom_point(aes(x=2,y=4),color="red",size=6,shape=63)
```

### IDW
We are now going to use IDW as a way to get an estimate of that point with a missing value of $z$. Since we are estimating it we call it $\hat{Z}$ and we know the location as $s_0$ so we call this $\hat{Z}(s_0)$. We'll follow the notation above and set the weight as a function of inverse distance with $p=2$.

The Euclidean distance to our unknown point from each of the known points is a vector: 
```{r,echo=FALSE}
d2s0 <- as.matrix(dist(cbind(c(foo$x,2),c(foo$y,4))))[1:5,6]
```

$$\mathbf{d_i} = \left[\begin{array}
{rr}
1 & 1.414\\
2 & 1.000\\
3 & 1.414\\
4 & 2.236\\
5 & 4.243
\end{array}\right]
$$

If we use a value of $p=2$ to get weights we have:

```{r,echo=FALSE}
p <- 2
w <- d2s0^-p
```

$$\mathbf{w(s_i)} = \left[\begin{array}
{rr}
1 & 0.500\\
2 & 1.000\\
3 & 0.500\\
4 & 0.200\\
5 & 0.056
\end{array}\right]
$$

Thus the interpolated point:
```{r,echo=FALSE}
zhat_s0 <- sum(w*foo$z)/sum(w)
```


$$\hat{Z}(s_0)=\frac{0.500(100)+1.000(105)+0.500(105)+0.200(100)+0.056(115)}{0.500+1.000+0.500+0.200+0.056} = 103.7$$
```{r, warning=FALSE}
# distance to missing point s0
d2s0 <- as.matrix(dist(cbind(c(foo$x,2),c(foo$y,4))))[1:5,6]
# power
p <- 2
# weights
w <- d2s0^-p
# and the estimation itself
zhat_s0 <- sum(w*foo$z)/sum(w)
zhat_s0
# add it to the plot
p1 + geom_point(aes(x=2,y=4,size=zhat_s0))
```

You can see how we interpolated (estimated) one point using inverse distance weighting. And you could imagine then how we'd do that for any point we could imagine.

::: {.callout-note}
If you need a refresher in `dist()`, work through the [Distance Matrices aside](03xAsideDistanceMatrices.html).
:::

## Application
### Data
Now that we've done a toy example, let's give this a shot with the lead data from the Meuse River. Because the lead data are pretty skewed, we will log transform it. First we'll load, then log the lead column, make it into a `sf` object and then plot.

First we will turn the `meuse` into class `sf`.

```{r}
data(meuse.all)
glimpse(meuse.all)
class(meuse.all)
meuse.all$logLead <- log(meuse.all$lead)
# or for the tidyverse fans this is the same output
meuse.all <- meuse.all %>% mutate(logLead = log(lead))
# make into sf
meuse_sf <- st_as_sf(meuse.all, coords = c("x", "y")) %>%
  st_set_crs(value = 28992)

class(meuse_sf) # note change in class from data.frame to sf and data.frame

p2 <- ggplot(data = meuse_sf) +
  geom_sf(aes(fill=logLead), size=4, 
          shape = 21, color="white",alpha=0.8)+
  scale_fill_continuous(type = "viridis",name="log(ppm)") + 
  labs(title="Lead concentrations")
p2
```

Let's interpolate those measurements to an empty grid that covers the whole study site. There is a handy object called `meuse.grid` that has a full grid of this area with 40-m cells stored as a `data.frame`. 

```{r}
meuse.grid <- readRDS("data/meuse.grid.Rds")
head(meuse.grid)
```

This has some of the fields (e.g., `soil`) that overlaps with `meuse`. We will use those later when we do some regression-based interpolation. Right now we just want the coordinates to predict onto which are in the first two columns. Let's make a `SpatRaster` object using `rast` form `terra`.


```{r}
meuse_grid_sf <- st_as_sf(meuse.grid, 
                          coords = c("x","y"), 
                          crs = st_crs(meuse_sf))
meuse_grid_sf
```

### IDW
So the idea here is that we will model a value for lead at each of the locations in `meuse_grid_sf`. Rather than do this by hand as we did above, we will use the very powerful `gstat` library.

We can use the coordinates in `meuse_grid_sf` as the locations to predict using IDW. We don't care about the data in `meuse_grid_sf` like soil type or distance to river. We just want the locations. To do this, we will use the `gstat` function which creates an object that holds all the information necessary for prediction. Here we tell `gstat` that we want an IDW model with a power of 2 using the `idp` parameter. The formula `logLead~1` tells `gstat` there are no covariates, just an intercept. The spatial weighting by distance is what makes this IDW -- that's handled by `idp`, not by the formula. This distinction will matter when we add covariates in regression kriging.

::: {.callout-note}
Note that we're calling `predict()` here which is a generic function that will dispatch to a different method under the hood because `idw_p2_model` is a `gstats` object. That's generics doing their job. Look back at the [Understanding Methods and Generics in R aside](01xAsideMethodsAndGenerics.html).
:::

```{r}
idw_p2_model <- gstat(formula=logLead~1, 
                      locations = meuse_sf,
                      set=list(idp = 2))
logLeadIDW_p2_sf <- predict(idw_p2_model,meuse_grid_sf)
logLeadIDW_p2_sf
```

Now, we are going to look at the interpolated data and assess if it's a "good" interpolation in a bit. But first, let's look at the `logLeadIDW_p2_sf` object we made. This is a `sf` object returned by `predict`. The interpolated lead levels are accessed via `logLeadIDW_p2_sf$var1.pred` (we will turn a blind eye to the `logLeadIDW_p2_sf$var1.var` column of NA values for *now* -- those NAs are there because IDW is deterministic; there's no probability model underneath it and therefore no variance to estimate. When we get to kriging, that column will actually mean something). All the `gstat` functions we are going to use follow this convention so the good news is that you only have to get used to it once.

But first, the `gstat` `predict` function returned a `sf` point object. We want it to be a `SpatRaster`. You'd think there would be an elegant way to convert an evenly spaced point object into a `SpatRaster` object but there isn't at the moment so we will brute force it. This is a little irksome but whatever. I'll write a little function to do this just because we'll have to do it a few times. If you haven't done this, writing functions is a good tool to have at your disposal.

```{r}
sf_2_rast <-function(sfObject,variableIndex = 1){
  # coerce sf to a data.frame
  dfObject <- data.frame(st_coordinates(sfObject),
                         z=as.data.frame(sfObject)[,variableIndex])
  # coerce data.frame to SpatRaster
  rastObject <- rast(dfObject,crs=crs(sfObject))
  
  names(rastObject) <- names(sfObject)[variableIndex]
  
  return(rastObject)
}

logLeadIDW_p2_rast <- sf_2_rast(logLeadIDW_p2_sf)
logLeadIDW_p2_rast
# and plot
ggplot() +
  geom_spatraster(data=logLeadIDW_p2_rast, mapping = aes(fill=var1.pred),alpha=0.8) +
  scale_fill_continuous(type = "viridis",name="log(ppm)",na.value = "transparent") + 
  labs(title="Lead concentrations", subtitle = "IDW with p=2") +
  theme_minimal()
```

We've made a prediction. Is it any good? It kind of looks funny -- especially all those craters in the middle. Let's take another look at it with a more extreme visualization.

```{r}
ggplot() +
  geom_spatraster_contour_filled(data=logLeadIDW_p2_rast,
                                 breaks = seq(from=3.5, to=6.5,by=0.25),
                                 alpha = 0.9) +
  scale_fill_discrete(name="log(ppm)",na.value = "transparent") + 
  labs(title="Lead concentrations", subtitle = "IDW with p=2") +
  theme_minimal()
```

Are those bumps and craters real? Or an artifact? Would the surface be better if we changed the value of $p$? What does "better" even mean? Would we know "better" if we saw it?

Those craters are artifacts of IDW, not real structure in the data. Here's why they happen: in areas where sample points are sparse, IDW has very little to work with. Any location in a gap between widely-spaced points ends up being pulled hard toward the one or two nearest observations, creating bullseye patterns around each isolated point. The center of the Meuse study area is sampled more thinly than the edges near the river, which is why that's where the craters are worst. No interpolation method handles sparse data well, but IDW is particularly vulnerable to this because its weighting scheme has no concept of the broader spatial structure of the data -- it just sees distance. Keep this in mind as we look at skill.

## Skill
Now that we are making predictions we need to assess how good they are. The [Model Skill and Cross Validation aside](05xAsideModelSkill.html) covers all of this in detail and is worth reading before going further. The short version is: we'll use R$^2$, RMSE, and the observed-vs-predicted plot throughout the interpolation chapters, and we always care more about out-of-sample skill than in-sample fit. Here's the in-sample picture first, just to show why:

```{r}
obs <- meuse_sf$logLead
preds <- extract(logLeadIDW_p2_rast, meuse_sf) %>% pull(var1.pred)
rsq <- cor(obs,preds)^2
rmse <- sqrt(mean((preds - obs)^2))
rsq
rmse
ggplot() +
  geom_abline(slope=1,intercept = 0) +
  geom_point(aes(x=obs,y=preds)) + 
  coord_cartesian() + 
  labs(x="Observed Values",
       y="Predicted Values",
       title="Lead log(ppm)")
```

The model does an excellent job of predicting the observed values with 99% variance explained and a low RMSE. It's 99% rather than 100% because we're extracting predictions from a raster grid, not from the exact observation coordinates -- there's a little rounding from the discretization. IDW is actually exact at data locations in theory (the weight to itself would be infinite, overwhelming everything else). Which is precisely why this in-sample comparison is kind of a lame test. The model knew about all those points when it made the predictions. Typically when assessing skill we withhold some portion of the data and then use that to test the model. When we test the model using the withheld data we are calculating the out-of-sample error by giving the model a chance to predict some data it hasn't seen before. E.g., let's withhold 25% of the data (38 points), do the interpolation, and then look at how well the model predicted the withheld points.

```{r}
n <- nrow(meuse_sf)
rows4test <- sample(x = 1:n,size = n*0.25)
meuseTest <- meuse_sf[rows4test,]
meuseTrain <- meuse_sf[-rows4test,]

# note that we build the model with meuseTrain
idw_p2_model <- gstat(formula=logLead~1, 
                      locations = meuseTrain,
                      set=list(idp = 2))
logLeadIDW_p2_sf <- predict(idw_p2_model,meuse_grid_sf)

logLeadIDW_p2_rast <- sf_2_rast(logLeadIDW_p2_sf)

# and plot
ggplot() +
  geom_spatraster(data=logLeadIDW_p2_rast, mapping = aes(fill=var1.pred),alpha=0.8) +
  scale_fill_continuous(type = "viridis",name="log(ppm)",na.value = "transparent") + 
  labs(title="Lead concentrations", subtitle = "IDW with p=2") +
  theme_minimal()
```

We built the model with the training data `meuseTrain` which had 75% of the original data. We kept 25% of the data hidden from the model in `meuseTest`. By comparing the predictions from the training data to the withheld values from the testing data we can assess the out-of-sample skill. 

```{r}
# note use of meuseTest here
obs <- meuseTest$logLead
preds <- extract(logLeadIDW_p2_rast, meuseTest) %>% pull(var1.pred)
rsq <- cor(obs,preds)^2
rmse <- sqrt(mean((preds - obs)^2))
rsq
rmse

ggplot() +
  geom_abline(slope=1,intercept = 0) +
  geom_point(aes(x=obs,y=preds)) + 
  coord_fixed(ratio=1, xlim = range(preds,obs),ylim = range(preds,obs)) +
  labs(x="Observed Values",
       y="Predicted Values",
       title="Lead log(ppm)")
```

Oooof. Looking at that plot and the code we see now that the skill of the model is greatly decreased with R$^2$ = `r round(rsq,3)` and RMSE = `r round(rmse,3)`. We can also see that the predicted values are less than the observed values for these points.

### Null Model

How does that RMSE look against the simplest possible baseline, predicting the training mean everywhere?

```{r}
rmseNULL <- sqrt(mean((mean(meuseTrain$logLead) - obs)^2))
rmseNULL
1 - (rmse / rmseNULL)
```

For a full treatment of the null model and how to interpret these skill scores, see the [Model Skill and Cross Validation aside](05xAsideModelSkill.html).

So we have to ask, "How can we improve this model in a responsible way?"

Improving a model's skill (without overfitting it) is an art. In this case, IDW is a pretty blunt tool. In fact it is completely **deterministic**. I.e., there is no probability theory underlying the method, which limits how much we can do.

What can we do right now? We could start trying to "tune" the model using parameters we can adjust. The power function is the easiest to mess with but we could also specify a minimum or maximum number of points used for the interpolation (arguments `nmax` and `nmin`) or a maximum distance that can influence the interpolation (`maxdist`).

For example, here is the same data as above with $p=3$ instead of $p=2$.

```{r}
idw_p3_model <- gstat(formula=logLead~1, 
                        locations = meuseTrain,
                        set=list(idp = 3))

logLeadIDW_p3_sf <- predict(idw_p3_model,meuse_grid_sf)

logLeadIDW_p3_rast <- sf_2_rast(logLeadIDW_p3_sf)

# and plot
ggplot() +
  geom_spatraster(data=logLeadIDW_p3_rast, mapping = aes(fill=var1.pred),alpha=0.8) +
  scale_fill_continuous(type = "viridis",name="log(ppm)",na.value = "transparent") + 
  labs(title="Lead concentrations") +
  theme_minimal()
```

How does this surface differ? Here is the skill on the withheld data:

```{r}
# note use of meuseTest here
obs <- meuseTest$logLead
preds <- extract(logLeadIDW_p3_rast, meuseTest) %>% pull(var1.pred)
rsq <- cor(obs,preds)^2
rmse <- sqrt(mean((preds - obs)^2))
rsq
rmse
ggplot() +
  geom_abline(slope=1,intercept = 0) +
  geom_point(aes(x=obs,y=preds)) + 
  coord_fixed(ratio=1, xlim = range(preds,obs),ylim = range(preds,obs)) +
  labs(x="Observed Values",
       y="Predicted Values",
       title="Lead log(ppm)")
```

The R$^2$ is higher and the RMSE is lower. It's a better fit but not much better. Against the same null, the skill score for $p=3$ is:

```{r}
1 - (rmse / rmseNULL)
```

That's a better model all around but not by much. Tuning $p$ only gets you so far with IDW -- the method just isn't well-suited to that sparse center of the study area, and no value of $p$ is going to fix thin data.

## Wrapping Up IDW

IDW is a reasonable starting point for spatial interpolation. It's fast, intuitive, and has no assumptions beyond Tobler's first law. But it is purely deterministic, so you get a surface but no uncertainty estimate. It's sensitive to the choice of $p$ but there's no principled way to pick it other than tuning by eye and cross-validation. And it has no way to account for spatial structure beyond distance: it can't know that lead decreases consistently with distance from the river, for instance. It just sees nearby points and weights them.

After a detour into thin-plate splines we'll tackle these problems. Kriging uses the variogram to model the spatial structure explicitly and, crucially, returns a variance estimate alongside the prediction. That `var1.var` column that was full of NAs? Kriging fills it in. After that, regression kriging incorporates covariates like distance to river directly into the model -- and that's where the formula `logLead~1` vs. `logLead~dist` distinction will make sense.

```{r, eval=FALSE, echo=FALSE}
#Optimizing p 3.058529

f = function(idp, formula, data,...){
  res <- sum(krige.cv(formula,data,set=list(debug=0,idp=idp),nfold=10,...)$residual**2)
  res
}
optimize(f, interval=c(0.01,4), formula=log(lead)~1, data=meuse_sf)
```

## Your Work
Let's turn to the California precipitation data that Hijmans uses for the nearest neighbor interpolation but we will do something cooler. We will make a continuous 10x10km surface of precipitation from the 432 locations of long-term precipitation records.

There are two files:

1. `prcpCA.rds` is a data frame with 432 points containing precipitation information. The columns `X` and `Y` give the coordinates of the station projected in California Albers [EPSG:3310](https://epsg.io/3310) with units in meters. We want to use the `ANNUAL` column which is total annual precipitation at that point in mm.

2. `gridCA.rds` is data frame with two columns `X` and `Y` that represent an empty grid of cells that we will interpolate the precip values into like we did with `meuse.grid` above. The spacing is 10x10 km for each point.

```{r}
# precip point data
prcpCA <- readRDS("data/prcpCA.rds")
# empty grid to interpolate into
gridCA <- readRDS("data/gridCA.rds")

# make as sf
prcpCA <- prcpCA %>% st_as_sf(coords = c("X", "Y")) %>%
  st_set_crs(value = 3310)

# simple map -- see postscript below for a fancy map
prcpCA %>% ggplot() + 
  geom_sf(aes(fill=ANNUAL,size=ANNUAL),color="white",
          shape=21,alpha=0.8) + 
  scale_fill_continuous(type = "viridis",name="mm") + 
  labs(title="Total Annual Precipitation") +
  scale_size(guide="none")
```

Use IDW to create a surface of total annual precipitation in California. Tune and assess skill as you see fit. Interpret and explain your surface and anything you notice in particular about it. A skill assessment of some kind is important so see the [Model Skill and Cross Validation aside](05xAsideModelSkill.html).

One warning. There are a few stations on the coast that are classified as `NA` in the gridded surface you'll make. This is because the gridded surface is 10x10 km and that coarse resolution means some of the edge cases (stations very close to the coast) get turned into "ocean" instead of "land". This is a really common issue when moving from point to raster data. It won't affect what you do until you try to assess skill. Then you'll find that some (eight, I think) of the stations will show up as `NA` when you use `extract`. This will cause your `rsq` and `rmse` calculations to return `NA`. You can overcome this in a variety of ways (e.g., `cor(obs,preds,use="complete.obs")^2` and `sqrt(mean((preds - obs)^2,na.rm=TRUE))` or with a mask etc.). You'd have figured that out on your own I bet, but I'm just giving you a heads up.


```{r,echo=FALSE, eval=FALSE}
gridCA <- gridCA %>% st_as_sf(coords = c("X", "Y")) %>%
  st_set_crs(value = 3310)

idw_p2.5_model <- gstat(formula=ANNUAL~1,locations=prcpCA,
                        set=list(idp = 2.5))

prcpIDW_p2.5_sf <- predict(idw_p2.5_model,gridCA)

prcpIDW_p2.5_rast <- sf_2_rast(prcpIDW_p2.5_sf)

# and plot
ggplot() +
  geom_spatraster(data=prcpIDW_p2.5_rast, mapping = aes(fill=var1.pred),alpha=0.8) +
  scale_fill_continuous(type = "viridis",name="mm",na.value = "transparent") + 
  labs(title="Total Annual Precip") +
  theme_minimal()


library(tmap)

tmap_mode("view")
tm_shape(prcpIDW_p2.5_rast) +
  tm_raster(col="var1.pred",alpha = 0.5) + #title = "TAP (mm)"
  tm_shape(prcpCA) + 
  tm_symbols(col="deeppink",alpha=0.5,size = "ANNUAL")

```

